\documentclass[a4paper]{llncs}

%\usepackage[T2A]{fontenc}
\usepackage[utf8x]{inputenc}
\usepackage[russian, english]{babel}
\usepackage{color}
\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,%
            citecolor=red,urlcolor=red]{hyperref}
\usepackage{graphicx}
\usepackage{listings}

\definecolor{lbcolor}{rgb}{0.95,0.95,0.95}
\newcommand{\codesize}{\fontsize{8}{8}\selectfont}
\lstset{% general command to set parameter(s)
    language=C,
    basicstyle=\codesize,       % print whole listing small
    %keywordstyle=\color{black}\bfseries,
                                % underlined bold black keywords
    identifierstyle=,           % nothing happens
    stringstyle=\ttfamily,      % typewriter type for strings
    showstringspaces=false%
}


\newcommand{\fixme}[1]{\vskip 5mm\noindent{\bf FIXME}: {\it #1}}

\author{Jūlija Pečerska\inst{1}, Artjoms Šinkarovs\inst{2}, Pavels Zaičenkovs\inst{3}}
\date{\today}
\title{On dynamic extensions of context-dependent parsers\\
       --- Extended Abstract ---}
\institute{
  University of Latvia,
  Raiņa bulvāris 19, Rīga,
  Latvija, LV-1586
\and
  Heriot-Watt University,
  Riccarton, Edinburgh,
  EH14 4AS, United Kingdom
\and
  Moscow Institute of Physics and Technology,
  141700, 9, Institutskii per., Dolgoprudny, 
  Moscow Region, Russia
}


\begin{document}
\maketitle

\begin{abstract}

Modern programming languages define a syntax that cannot be described precisely
using context free grammars.  This problem is well known, and as a solution
there are parser generators like ANTLR, which generates LL(*) parsers rather
than LALR/LR.  Nevertheless, a lot of real-world parsers are implemented by
hand.  A great difficulty arises when there appears a need to introduce some
changes to the grammar of the language.  This implies that the parser
has to be rewritten to include new rules and resolve the created
ambiguities manually.

Whenever radical changes to the language are required, it is understood that
parser rewriting will be necessary. However, when syntax requires some
extensions to the convenience of the user, it is natural to allow her to adapt
the existing syntax of the selected programming language to her needs.  In this
paper we present a way of building a system that allows introducing dynamic
changes to the grammar and that works on top of any recursive descent parser
that meets some minor restrictions.

The execution of the proposed system consists of 2 steps. First is matching a
list of tokens against a regular expression built on tokens and encoded
production names of the given grammar. And the second is transforming the
result using a functional language.  We demonstrate that using this approach
makes it possible to i) give static guarantees regarding the transformation 
rules; ii) safely express non-trivial syntactical constructions; and iii)
perform a restricted partial evaluation.
\end{abstract}

\input{intro.tex}

\input{parser-model.tex}

\input{dynamic-extensions.tex}

\input{regexps.tex}

\input{application.tex}

\input{conclusion.tex}
%\section{Evaluation}
%Here is a bunch of links for the existing macro-preprocessors:
%
%\begin{tabular}{l | p{.8\textwidth}}
%ML/I & \url{http://www.ml1.org.uk/htmldoc/ml1sig.html} \\
%GEMA & \url{http://gema.sourceforge.net/new/docs.shtml} \\
%GPP  & \url{http://files.nothingisreal.com/software/gpp/gpp.html}
%В этой штуке советую заглянуть в ADVANCED EXAMPLES с лямбдой\\
%TRAC &
%\url{http://web.archive.org/web/20050205172849/http://tracfoundation.org/t2001tech.htm}
%Это очень разумная идея правда совсем дохлая -- там тоже
%функциональный язык внутри живет, но работает на строках кажись
%\end{tabular}
%
%Еще бывают: m4, cpp, lisp/scheme macros, tex?...
%\input{lisp-macros.tex}
%\input{mli-macros.tex}
%\input{cpp-templates.tex}
%
%\section{Future work}

\bibliographystyle{plain}
\bibliography{paper}


\end{document}
